# Reproducible Research: Peer Assessment 1
## author: Matthew A. Lanham


The dataset description:
- The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
1. steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
2. date: The date on which the measurement was taken in YYYY-MM-DD format
3. interval: Identifier for the 5-minute interval in which measurement was taken

```{r, echo=TRUE}
## Loading and preprocessing the data
data = read.csv("activity\\activity.csv")
head(data)
```

## What is mean total number of steps taken per day?
```{r, echo=TRUE, fig.width=10, fig.height=5}
## Create a new data set calculating the total, mean, and median steps per day
my_dates = as.data.frame(table(data$date))[,1]
data2 = data
data2$date = as.ordered(data2$date)
listOfSums = by(data2[,c("steps","interval")], data2$date, colSums)
listOfMeans = by(data2[,c("steps","interval")], data2$date, colMeans)
require(miscTools) #needed for colMedians() function
listOfMedians = by(data2[,c("steps","interval")], data2$date, colMedians)
summaryByDay = as.data.frame(matrix(nrow=length(my_dates), ncol=4))
colnames(summaryByDay) = c("date","total","mean","median")
for (i in 1:length(my_dates)) {
    summaryByDay[i,1] = as.character(my_dates[i])          #date
    summaryByDay[i,2] = listOfSums[[my_dates[i]]][[1]]     #total
    summaryByDay[i,3] = listOfMeans[[my_dates[i]]][[1]]    #mean
    summaryByDay[i,4] = listOfMedians[[my_dates[i]]][[1]]  #median
}

## Create a Histogram
hist(summaryByDay[,2], col="lightblue"
     , main="Histogram of total number of steps per day"
     , xlab="Total steps"
     , ylab="Frequency")

## Report of mean and median total number of steps by day (ignore missing values)
summaryByDay[complete.cases(summaryByDay),c("date","mean","median")]
```

## What is the average daily activity pattern?
```{r, echo=TRUE, fig.width=10, fig.height=5}
## Create a dataset showing the mean per interval for the time-series plot
my_intervals = as.data.frame(table(data$interval))[,1]
data3 = data
data3$interval = as.ordered(data3$interval)
data3$date = as.numeric(data3$date)
listOfIntervalMeans = by(data3[,c("steps","date")], data3$interval, colMeans, na.rm=TRUE)
summaryByInterval = as.data.frame(matrix(nrow=length(my_intervals), ncol=2))
colnames(summaryByInterval) = c("interval","mean")
for (i in 1:length(my_intervals)) {
    summaryByInterval[i,1] = as.character(my_intervals[i])                #interval
    summaryByInterval[i,2] = listOfIntervalMeans[[my_intervals[i]]][[1]]  #mean
}
summaryByInterval

## time-series plot of mean steps per 5-minute interval
plot(summaryByInterval$interval, summaryByInterval$mean, type="l"
     , main="Average number of steps per interval"
     , xlab="5-minute time interval"
     , ylab="Mean number of steps"
     , col="blue", lwd=2
     )

## interval containing max number of steps
summaryByInterval[which.max(summaryByInterval$mean),]
```
## Imputing missing values
```{r, echo=TRUE, fig.width=10, fig.height=5}
data4 = data
## report total number of missing values in the dataset
dim(data4[!complete.cases(data4),])[1]

## Here I imputed the missing values using mean for each time interval
for (i in 1:dim(summaryByInterval)[1]) {
    #identify missing locations within each interval
    missingSet = !complete.cases(data4[which(data4$interval==summaryByInterval[i,1]),])
    #set those missing values within each interval to its corresponding mean
    data4[which(data4$interval==summaryByInterval[i,1]),][missingSet,"steps"] = summaryByInterval[i,2]
}

## Create a new data with the new imputed values
data4$date = as.ordered(data4$date)
listOfSumsImp = by(data4[,c("steps","interval")], data4$date, colSums)
listOfMeansImp = by(data4[,c("steps","interval")], data4$date, colMeans)
require(miscTools) #needed for colMedians() function
listOfMediansImp = by(data4[,c("steps","interval")], data4$date, colMedians)
summaryByDayImp = as.data.frame(matrix(nrow=length(my_dates), ncol=4))
colnames(summaryByDayImp) = c("date","total","mean","median")
for (i in 1:length(my_dates)) {
    summaryByDayImp[i,1] = as.character(my_dates[i])             #date
    summaryByDayImp[i,2] = listOfSumsImp[[my_dates[i]]][[1]]     #total
    summaryByDayImp[i,3] = listOfMeansImp[[my_dates[i]]][[1]]    #mean
    summaryByDayImp[i,4] = listOfMediansImp[[my_dates[i]]][[1]]  #median
}

## Create a Histogram of the steps per day AFTER imputation
hist(summaryByDayImp[,2], col="lightblue"
     , main="Histogram of total number of steps per day"
     , xlab="Total steps"
     , ylab="Frequency")

## Report of mean and median total number of steps by day (with imputed values)
summaryByDayImp[complete.cases(summaryByDayImp),c("date","mean","median")]

## Side-by-side comparision by day after imputation
results = merge(x=summaryByDay, y=summaryByDayImp, by="date")
results$date = as.Date(results$date)
names(results) = c("date","total","mean","median","totalImp","meanImp","medianImp")
results

## The difference in total step after imputation for days that didn't have missing values initially
resultsClean = results[complete.cases(results),]
summary(resultsClean$totalImp - resultsClean$total)

## Addttional steps over 60-day window because of imputation
sum(results$totalImp) - sum(results$total, na.rm=TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE, fig.width=10, fig.height=10}
## Create a new variable called "dayInd" showing if the measurement was taken during the weekend or weekday
data4$date = as.Date(data4$date)
require(timeDate)
data4$dayInd = as.factor(ifelse(isWeekday(data4$date),"weekday","weekend"))
my_intervals = as.data.frame(table(data$interval))[,1]
data4$interval = as.ordered(data4$interval)
data4$date = as.numeric(data4$date)
weekend = data4[which(data4$dayInd=="weekend"),]
weekday = data4[which(data4$dayInd=="weekday"),]

## Generate the stats that will be used for the time-series plot below
listOfWeekendMeans = by(weekend[,c("steps","date")], weekend$interval, colMeans, na.rm=TRUE)
listOfWeekdayMeans = by(weekday[,c("steps","date")], weekday$interval, colMeans, na.rm=TRUE)
weekendSummary = as.data.frame(matrix(nrow=length(my_intervals), ncol=2))
weekdaySummary = as.data.frame(matrix(nrow=length(my_intervals), ncol=2))
colnames(weekendSummary) = c("interval","mean")
colnames(weekdaySummary) = c("interval","mean")
for (i in 1:length(my_intervals)) {
    weekendSummary[i,1] = as.character(my_intervals[i])                #interval
    weekendSummary[i,2] = listOfWeekendMeans[[my_intervals[i]]][[1]]   #mean
    weekdaySummary[i,1] = as.character(my_intervals[i])                #interval
    weekdaySummary[i,2] = listOfWeekdayMeans[[my_intervals[i]]][[1]]   #mean
}

## time-series plot of mean steps per 5-minute interval
par(mfrow=c(2,1))
plot(weekdaySummary$interval, weekdaySummary$mean, type="l"
     , main="Average number of steps per interval (Weekday)"
     , xlab="5-minute time interval"
     , ylab="Mean number of steps"
     , col="blue", lwd=2
     )
plot(weekendSummary$interval, weekendSummary$mean, type="l"
     , main="Average number of steps per interval (Weekend)"
     , xlab="5-minute time interval"
     , ylab="Mean number of steps"
     , col="blue", lwd=2
     )
```